\chapter{Background}

for each trajectory $j = 1, ..., J$ we define 
the SDE of the log-price $X_t^j$ as:
\begin{equation}
    \label{log-price SDE}
    \begin{aligned}
       &dX_t^j = -\frac{1}{2}\sigma_j^2 dt + \sigma_j dW_t\\
         &\text{with } X_0^j = \log(S_0^j)
    \end{aligned}
    \end{equation} 

Where $S_0^j$ is the raw price at time $0$ of trajectory $j$, $\sigma_j$ is the volatility of trajectory $j$ and $W_t$ is a standard Brownian motion.\\
$x_0^j$ and $\sigma_j$ are randomly sampled for each trajectory such that 
$x_0^j \in [0, 10]$ and $\sigma_j \in (0.0, 1.0]$.\\

In this context our condition label $y_j$ for a general data observation $j$ is given by
a vector containing the trajectory $y_j=[X_0^j, X_{dt}^j, X_{2dt}^j, ..., X_{Ndt}^j]$.
So our condition data matrix $Y$ has shape $J\times N$ where $J$ is the number of trajectories (data points)
 and $N$ is the number of time steps in each trajectory.\\
 
 The SDE is discretized over a uniform time space $N dt = t$ obtaining the following form
 adopted for the simulation of the trajectories:
 \begin{equation}
    X_{t+dt}^j = X_0^j -\frac{1}{2}\sigma_j^2 dt + \sigma_j \sqrt{dt} Z_t
 \end{equation}
 Where $Z_t \sim \mathcal{N}(0,1)$.\\
The procedure to build the condition data matrix $Y$ is summarized in Algorithm \ref{alg:y condition}.

\begin{algorithm}[H]
\caption{Algorithm for trajectory simulation and condition data matrix $Y$ creation}
\label{alg:y condition}
\begin{algorithmic}[1]

\STATE Define the number of simulation $J$.
\STATE Initialize an empty data matrix $Y$ of shape $J\times N$.
\STATE Set iteration counter $j \leftarrow 0$.
\REPEAT
    \STATE Randomly sample $X_0^j \in [0.0, 10.0]$ and $\sigma_j \in (0.0, 1.0]$.
    \STATE Evolve $X_t^j$ based on the discretized SDE for $0, dt, ..., Ndt$ steps.
    \STATE Store the trajectory $[X_0^j, X_{dt}^j, X_{2dt}^j, ..., X_{Ndt}^j]$ in the $j$-th row of $Y$.
    \STATE Increment counter $j \leftarrow j + 1$.
\UNTIL{$j = J$}
\RETURN $Y$
\end{algorithmic}
\end{algorithm}



\section{Distribution parameters as input}
In $t=N dt$ we have the following information: the final log-price $X_{Ndt}^j$ of each trajectory $j$ and the corresponding volatility $\sigma_j$.
The goal of the cGAN is to learn the distribution of the next time interval $Ndt + \Delta t$ 
which is a normal distribution $\mathcal{N}(X_{Ndt}^j - \frac{1}{2}\sigma_j^2 \Delta t, \sigma_j \sqrt{\Delta t})$.\\
So the data matrix $X$, that in classical implementations of GANs are images, is in this case 
a vector containing the parametrization of the distribution. In this case a vector of two elements:
the mean and the standard deviation.

\section{Approximated distribution via bins as input}
An alternative approach to represent the distribution at time $Ndt + \Delta t$ is to approximate it via bins.
In this case the data matrix $X$ is a vector of dimension $B$ where $B$ is the number of bins ranges adopted for the approximation.
Each element of the vector represents the expected value given that the value falls inside that bin range, so 
$E[ X_{Ndt}^j | b_i \leq X_{Ndt}^j \leq b_{i+1}]$ for $X_{Ndt}^j \sim \mathcal{N}(X_{Ndt}^j - \frac{1}{2}\sigma_j^2 \Delta t, \sigma_j \sqrt{\Delta t})$, in the following description
we just define the mean parameter as $\mu_{X^j}$ and the standard deviation parameter as ${\sigma_{X^j}}$.
To create this representation we first define the bin edges by taking as min and max values the $ \pm 4 \sigma_{X^j}$ values coming from the normal distribution previously defined.
Then we create $B$ bins uniformly spaced between the min and max values.
For each bin range $B_i$ defined by the edges $[b_{i}, b_{i+1}]$,
we compute the expected value for each bin by using the following formula:
\begin{equation}
    \mathbb{E}[Z \mid a_i \leq Z \leq a_{i+1}] = \frac{\phi(a) - \phi(a_{i+1})}{\Phi(a) - \Phi(a_{i+1})}
\end{equation}
where $a_{i}$ and $a_{i+1}$ are the standardized bins, $\phi(z)=\frac{1}{\sqrt{2\pi}} e^{-z^2/2}$ is the standard normal probability density function and 
$\Phi(z) = \int_{-\infty}^{z} \phi(t) dt$. Finally, the expected value of $Z$ inside the two standardized bins is converted back to its original scale by multiplying it by ${\sigma_{X^j}}$ and adding $\mu_{X^j}$.



\subsection{The Kolmogorov-Smirnov test}
The Kolmogorov-Smirnov (KS) goodness-of-fit test evaluates whether an empirical distribution is consistent with a fully specified continuous reference distribution. 
Let $F(x)$ denote the hypothesized cumulative distribution function (CDF), 
and let $F_{n}(x)$ be the empirical CDF constructed from a sample of size $n$. 
The KS statistic is defined as:
\begin{equation}
    D_{n}=\sup_{x}\lvert F_{n}(x)-F(x)\rvert
\end{equation}
representing the maximum vertical deviation between the empirical and theoretical CDFs.
Under the null hypothesis $H_{0}: F_{n}(x)=F(x)$
the scaled distribution $\sqrt{n}D_{n}$ converges to the Kolmogorov distribution, enabling computation of p-values via:
\begin{equation*}
    P(\sqrt{n}D_{n}\le t)=1-2\sum_{k=1}^{\infty}(-1)^{k-1}e^{-2k^{2}t^{2}}
\end{equation*}
Large values of $D_{n}$ lead to rejection of $H_{0}$, indicating
incompatibility between the empirical and theoretical distributions.\\

When applying the KS framework to generated outputs from the proposed GAN, the empirical object is a vector of bin probabilities
$(p_{1},\dots,p_{K})$  These represent a discretized approximation of the model’s PDF. To construct the empirical CDF, 
define cumulative probabilities:
\begin{equation*}
    F_{\text{GAN}}(x_{k})=\sum_{i=1}^{k}p_{i}
\end{equation*}

where $x_k$ indexes the bin boundaries. For comparison with a normal distribution, compute the theoretical CDF
$F_{\mathcal N}(x_{k})$ at the same grid. The KS statistic becomes:
\begin{equation*}
    D=\max_{k}\lvert F_{\text{GAN}}(x_{k})-F_{\mathcal N}(x_{k})\rvert
\end{equation*}

This yields a valid discrepancy measure, although interpreting p-values requires specifying an effective sample size, since the GAN vector itself is not an empirical sample.
Nonetheless, $D$ remains a rigorous, distribution-free indicator of alignment between the GAN-generated distribution and the target normal CDF.
We can actually compute the the p-value of the test using the survival function of the kolmogorov distribution since $P(D_n ≥ \hat{D_n} | H_0)$.

\section{CGAN formulation}
GANs are often formalized as a minimax game, where the generator (G) seeks to minimize the value function V while the discriminator (D) aims to maximize it:
\begin{equation}
    \label{eq:GAN minimax}
\min _G \max _D V(D, G)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

The discriminator outputs a value $D(x)$, 
which represents the probability that $x$ is the real vector containing mean and variance of the normal distribution.
Its goal is to correctly classify the real parameters as real and generated parameters as fake,
effectively maximizing the likelihood of the observed data.
To measure this, we use cross-entropy loss, as is common in deep learning,
expressed as $plog(1-p)$. 
For real parameters, the true label $p$ equals 1, 
while for generated parameters the label is 0.
With this setup, the discriminator's objective becomes:
\begin{equation}
    \label{eq:discriminator objective}
    \max _D V(D)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

where $\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]$ define how good the discriminator recognize real parameters.
and $\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]$ define how good the discriminator recognizes generated parameters.\\
The generator's objective is to produce distribution parameters that maximize $D(x)$, pushing the discriminator to classify them as real:
\begin{equation}
    \label{eq:generator objective}
\min _G V(G)=\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

Where $\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]$ define how good the generator is in fooling the discriminator.\\

In the Conditional GAN (CGAN), the generator learns to generate a fake sample with a specific condition, in our case the past trajectory, rather than a generic sample from unknown noise distribution.
The CGAN objective function is given by:
\begin{equation}
    \label{eq:CGAN minimax}
    \min _G \max _D V(D, G)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x} \mid \boldsymbol{y})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z} \mid \boldsymbol{y})))]
\end{equation}

Academic findings highlighted in \cite{background:gan_time_series} suggest that one of the principal failure modes of GANs 
arises when the generator converges to a parameter configuration that yields only a limited variety of outputs.
Existing research indicates that the occurrence of mode collapse is attributed to the establishment of a local 
equilibrium within the minimax optimization framework.
A further critical challenge involves the vanishing gradient problem,
which emerges when the discriminator becomes overly proficient,
causing the generator's gradients to approach zero and consequently impeding its learning process.



\section{Possible improvement to CGAN}
Different strategies have been adopted to improve training stability and convergence 
of GANs.
\begin{itemize}
    \item \textbf{Wasserstein GAN}: WGAN \cite{background:wgan} introduces the Wasserstein distance (Earth Mover's Distance) as a new loss function for GANs. This distance metric provides a more meaningful measure of the difference between the real and generated data distributions,
     leading to improved training stability and convergence. The WGAN objective function is given by $\min _G \max _{D \in \mathcal{D}} E_{\boldsymbol{x} \sim p_d}[D(\boldsymbol{x})]-E_{\widehat{\boldsymbol{x}} \sim p_g}[D(\widehat{\boldsymbol{x}})]$.
     \item \textbf{DRAGAN}: this framework was introduced to address these challenges by incorporating a gradient penalty directly into the GAN architecture.
    The authors in \cite{background:dragan} observed that the local equilibria frequently manifest steep discriminator gradients in the vicinity of certain real data samples.
    They further demonstrated that such degenerate local equilibria can be mitigated by applying a gradient penalty to the objective function, the method adopted is called DRAGAN.
    \item A tip suggested in \cite{background:gan_time_series} to do not use batch normalization after adding the gradient penalty in order to improve model performances.
    \item in \cite{background:rcgan} both the generator and discriminator are RNN or LSTM networks adopted to better capture the sequential dependencies in time series data.
\end{itemize}


