\chapter{Introduction}

The objective of this thesis is to design and implement a Conditional Generative Adversarial Network (CGAN) which is able to generate 
the future probability distribution of the price of an asset given its past trajectory.\\
In particular, the Generator receives as input a noise vector and a sequence of past prices observations where the price dynamics is modeled
as a Geometric Brownian Motion. The output of the Generator is a vector representing the discretized probability density function of the future price of the asset at a given time horizon.\\

\section{Training Data}
In order to effectively train the CGAN model, two kinds of data are required for the training phase:
\begin{itemize}
    \item Condition input data $C$: it is a $J \times N$ matrix, where $J$ is the number of trajectories and $N$ is the number of time steps of each trajectory.
     This data represents the past $J$ trajectories of the asset prices, where each trajectory is associated with a specific volatility parameter $\sigma_j$ and eventually a specific starting point $X_0^j$, and is used to condition both the Generator and the Discriminator.
    \item Target output data $X$: it is a $J \times B$ matrix, where $J$ represents the number of trajectories and $B$ represents the number of elements used to describe the probability density function. 
    This data represents the true theoretical future probability distributions of the asset prices associated to their respective past trajectories for a given time horizon. 
    It is used to train the Discriminator to distinguish between real and fake data.
\end{itemize}


\subsection{Condition input data}
For each trajectory $j = 1, ..., J$ we define 
the SDE of the log-price $X_t^j$ as:
\begin{equation}
    \label{log-price SDE}
    \begin{aligned}
       &dX_t^j = -\frac{1}{2}\sigma_j^2 dt + \sigma_j dW_t\\
         &\text{with } X_0^j = \log(S_0^j)
    \end{aligned}
    \end{equation} 

Where $S_0^j$ is the raw price at time $0$ of trajectory $j$, $\sigma_j$ is the volatility of trajectory $j$ and $W_t$ is a standard Brownian motion.\\
$X_0^j$ and $\sigma_j$ are randomly sampled for each trajectory such that 
$X_0^j \in [-1, 1]$ and $\sigma_j \in (0.0, 0.5]$.\\

In this context our condition $c_j$ for a general data observation $j$ is given by
a vector containing the trajectory $c_j=[X_0^j, X_{dt}^j, X_{2dt}^j, ..., X_{Ndt}^j]$.

 The SDE is discretized over a uniform time space $N dt = T$ obtaining the following form
 adopted for the simulation of the trajectories:
 \begin{equation}
    X_{t+dt}^j = X_t^j -\frac{1}{2}\sigma_j^2 dt + \sigma_j \sqrt{dt} Z_t
 \end{equation}
 Where $Z_t \sim \mathcal{N}(0,1)$.\\
The procedure to build the condition data matrix $C$ is summarized in Algorithm \ref{alg:y condition}.

\begin{algorithm}[H]
\caption{Algorithm for the condition data matrix $C$ creation}
\label{alg:y condition}
\begin{algorithmic}[1]

\STATE Define the number of simulation $J$.
\STATE Initialize an empty data matrix $C$ of shape $J\times N$.
\STATE Set iteration counter $j \leftarrow 0$.
\REPEAT
    \STATE Randomly sample $X_0^j \in [-1.0, 1.0]$ and $\sigma_j \in (0.0, 0.5]$.
    \STATE Evolve $X_t^j$ based on the discretized SDE for $0, dt, ..., Ndt$ steps.
    \STATE Store the trajectory $[X_0^j, X_{dt}^j, X_{2dt}^j, ..., X_{Ndt}^j]$ in the $j$-th row of $Y$.
    \STATE Increment counter $j \leftarrow j + 1$.
\UNTIL{$j = J$}
\RETURN $Y$
\end{algorithmic}
\end{algorithm}


\subsection{Target output data}


At the end of the condition trajectory, so at time $T=N dt$, we have the following information: the final log-price $X_{Ndt}^j$ of each trajectory $j$ and the corresponding volatility $\sigma_j$.
The goal of the Conditional GAN is to learn the distribution of final values of the stochastic process at next time horizon $T + \tau$. 
A theoretical closed form can be used as ground truth. It is given by a normal distribution:
\begin{equation}
    \begin{aligned}
    X_{Ndt + \tau}^j & \quad \sim \quad \mathcal{N}(X_{Ndt}^j - \frac{1}{2}\sigma_j^2 \Delta t, \sigma_j \sqrt{\Delta t})\\
    \text{where } &\Delta t = dt \times \tau \text{ with $\tau$ the number of days ahead. }
    \end{aligned}
\end{equation}

So the output $x_j$ which has to be generated by the Generator is a representation of this distribution.
Two different approaches are considered to represent this distribution:
\begin{itemize}
    \item Using the parameters of the normal distribution as output: This represents the simplest way to represent the distribution since only two parameters are needed which are characterized by a closed form.\\
    In fact, in this case the output data matrix $X$ is a $J \times 2$ matrix where the first column contains the means $\mu_X = X_{Ndt} - \frac{1}{2}\sigma^2 \Delta t$
     and the second column contains the standard deviations $\sigma_X = \sigma \sqrt{\Delta t}$ of the normal distributions. It is important to note that in order to compute these parameters the final log-price $X_{Ndt}$ of the past trajectory should be known.
    \item Using a discretized normal distribution as output: In this case the output data matrix $X$ is a $J \times 100$ since for each trajectory the future distribution is discretized into 100 bins.
    The steps needed to obtain the following representation are the following:
    \begin{algorithm}[H]
    \caption{Algorithm for the target data matrix $X$ creation}
    \label{alg:x target}
    \begin{algorithmic}[1]

    \STATE Compute $J$ trajectories using algorithm \ref{alg:y condition}.
    \STATE Compute $\mu_X = X_{Ndt} - \frac{1}{2}\sigma^2 \Delta t$ and $\sigma_X = \sigma \sqrt{\Delta t}$.
    \STATE Compute the data matrices $X_{lower} = \mu_X - 4\sigma_X$ and $X_{upper} = \mu_X + 4\sigma_X$.
    \STATE Compute the global minimum $X_{min} = \min(X_{lower})$ and maximum $X_{max} = \max(X_{upper})$.
    \STATE Create $101$ evenly spaced bins edges between $X_{min}$ and $X_{max}$. This representation is shared by each sample, and it is stored into a vector $b$ of dimension $101$.
    \STATE Initialize an empty data matrix $X$ of shape $J\times 100$.
    \STATE Set iteration counter $j \leftarrow 0$.
    \REPEAT
        \STATE Standardize the bins edges in $b$ using $\mu_X^j$ and $\sigma_X^j$ to obtain the standardized bin edges $s_i^j = \frac{b_i - \mu_X^j}{\sigma_X^j}$ for $i = 1, ..., 101$.
        \STATE Apply the standard normal cumulative distribution function $F_Z$ to each standardized bin edge $s_i^j$.
        \STATE Compute the difference between two consecutive bin edges $p_i^j = F_Z(s_{i+1}^j) - F_Z(s_i^j)$ to obtain the density inside each bin range.
        \STATE Store the resulting vector of dimension $100$ in the $j$-th row of $X$.
        \STATE Increment counter $j \leftarrow j + 1$.
    \UNTIL{$j = J$}
    \RETURN $X$
    \end{algorithmic}
    \end{algorithm}
\end{itemize}



\section{Models and tests overview}
In the following section the architecture of proposed Conditional GAN will be described in addition with some architectural variations.
Moreover, the Kolmogorov-Smirnov test will be presented in order to evaluate the performance of the model in generating the future probability distribution of the asset price.\\

\subsection{CGAN formulation}
GANs are often formalized as a minimax game, where the generator (G) seeks to minimize the value function V while the discriminator (D) aims to maximize it:
\begin{equation}
    \label{eq:GAN minimax}
\min _G \max _D V(D, G)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

The discriminator outputs a value $D(x)$, 
which represents the probability that $x$ is the real vector containing mean and variance of the normal distribution.
Its goal is to correctly classify the real parameters as real and generated parameters as fake,
effectively maximizing the likelihood of the observed data.
To measure this, we use cross-entropy loss, as is common in deep learning,
expressed as $plog(1-p)$. 
For real parameters, the true label $p$ equals 1, 
while for generated parameters the label is 0.
With this setup, the discriminator's objective becomes:
\begin{equation}
    \label{eq:discriminator objective}
    \max _D V(D)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

where $\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x})]$ define how good the discriminator recognize real parameters.
and $\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]$ define how good the discriminator recognizes generated parameters.\\
The generator's objective is to produce distribution parameters that maximize $D(x)$, pushing the discriminator to classify them as real:
\begin{equation}
    \label{eq:generator objective}
\min _G V(G)=\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]
\end{equation}

Where $\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z})))]$ define how good the generator is in fooling the discriminator.\\

In the Conditional GAN (CGAN), the generator learns to generate a fake sample with a specific condition, in our case the past trajectory, rather than a generic sample from unknown noise distribution.
The CGAN objective function is given by:
\begin{equation}
    \label{eq:CGAN minimax}
    \min _G \max _D V(D, G)=\mathbb{E}_{\boldsymbol{x} \sim p_{\text {data }}(\boldsymbol{x})}[\log D(\boldsymbol{x} \mid \boldsymbol{y})]+\mathbb{E}_{\boldsymbol{z} \sim p_{\boldsymbol{z}}(\boldsymbol{z})}[\log (1-D(G(\boldsymbol{z} \mid \boldsymbol{y})))]
\end{equation}

Academic findings highlighted in \cite{background:gan_time_series} suggest that one of the principal failure modes of GANs 
arises when the generator converges to a parameter configuration that yields only a limited variety of outputs.
Existing research indicates that the occurrence of mode collapse is attributed to the establishment of a local 
equilibrium within the minimax optimization framework.
A further critical challenge involves the vanishing gradient problem,
which emerges when the discriminator becomes overly proficient,
causing the generator's gradients to approach zero and consequently impeding its learning process.

\subsection{Wasserstein CGAN formulation}

In the following section the Conditional Wasserstein Generative Adversarial Network (CWGAN) architecture is introduced.
It extends the traditional conditional GAN framework by means of the Wasserstein distance exploitation.\\

The main changes that characterize the CWGAN with respect to the CGAN are:
\begin{itemize}
    \item The loss function of a standard GAN is based on Jensen-Shannon divergence minimization while the CWGAN is based on the Wasserstein distance, which provides a more stable training process and mitigates issues such as mode collapse.
    \item The discriminator, which in Wasserstein GAN architecture is called critic, is designed to output real-valued scores instead of probabilities, allowing it to better capture the differences between real and generated data distributions.
    For this reason, the discriminator network does not use a sigmoid activation function in the output layer.
    \item The training procedure requires as a good practice to optimize the critic multiple times per generator update to ensure that the critic provides accurate gradients for the generator.
    \item The use of gradient penalty term in the loss function has the objective to enforce the Lipschitz constraint required for the Wasserstein distance computation.  
\end{itemize}


Now the critic maximizes the following loss function $E_{\boldsymbol{x} \sim p_d}[D(\boldsymbol{x})]-E_{\widehat{\boldsymbol{x}} \sim p_g}[D(\widehat{\boldsymbol{x}})]$ where $\boldsymbol{x}$ are real data samples from the data distribution $p_d$ and $\widehat{\boldsymbol{x}}$ are generated samples from the generator distribution $p_g$.
On the other hand, the generator minimizes the loss function $-E_{\widehat{\boldsymbol{x}} \sim p_g}[D(\widehat{\boldsymbol{x}})]$.\\
Given this setup, there is no need anymore for true and fake data labels since the critic outputs real-valued scores.\\
An additional advatage of using the Wasserstein distance is that it provides a meaningful measure of the distance between the real and generated data distributions, which allows for the implementation of an early stopping technique which
improve the training efficiency.\\


\subsection{The Kolmogorov-Smirnov test}
The Kolmogorov-Smirnov (KS) as defined in \cite{background:ks_test}, is a goodness of fit test that evaluates whether an empirical distribution is consistent with a fully specified continuous reference distribution. 
Let $F(x)$ denote the hypothesized cumulative distribution function (CDF), 
and let $F_{n}(x)$ be the empirical CDF constructed from a sample of size $n$. 
The KS statistic is defined as:
\begin{equation}
    D_{n}=\sup_{x}\lvert F_{n}(x)-F(x)\rvert
\end{equation}
representing the maximum vertical deviation between the empirical and theoretical CDFs.
Under the null hypothesis $H_{0}: F_{n}(x)=F(x)$,
the scaled distribution $\sqrt{n}D_{n}$ converges to the Kolmogorov distribution, enabling computation of p-values by using its CDF:
\begin{equation*}
    P(\sqrt{n}D_{n}\le t)=1-2\sum_{k=1}^{\infty}(-1)^{k-1}e^{-2k^{2}t^{2}}
\end{equation*}
Large values of $D_{n}$ lead to rejection of $H_{0}$, indicating
incompatibility between the empirical and theoretical distributions.\\

When applying the KS framework to generated outputs from the proposed GAN, the empirical object is a vector of bin probabilities
$(p_{1},\dots,p_{K})$.  These represent a discretized approximation of the model's PDF. To construct the empirical CDF, 
the cumulative probabilities are computed as:
\begin{equation*}
    F_{\text{GAN}}(x_{k})=\sum_{i=1}^{k}p_{i}
\end{equation*}

where $x_k$ indexes the bin boundaries. For comparison with a normal distribution, the theoretical CDF
$F_{\mathcal N}(x_{k})$ is computed at the same grid. The KS statistic becomes:
\begin{equation*}
    D=\max_{k}\lvert F_{\text{GAN}}(x_{k})-F_{\mathcal N}(x_{k})\rvert
\end{equation*}

This yields a valid discrepancy measure, although interpreting p-values requires specifying an effective sample size.
Nonetheless, $D$ remains a rigorous, distribution-free indicator of alignment between the GAN-generated distribution and the target normal CDF.
Actually, in order to compute the p\-value of the test the survival function of the Kolmogorov distribution is used $P(\sqrt{n}D_n \geq \sqrt{n}\hat{D_n} | H_0)$ 
where $\hat{D_n}$ is the observed test statistic. 
The effective sample size $n$ has been set equal to the number of bins $B$ used to discretize the distribution.

